CSD, Disjoint components and trayectory detection

### 2020-11-14 ArticuloCSDyCM/Paper03.tex
simplifique oraciones

Tracking the activity in neural tissue poses challenging problems, if one
wants a clear, spatial and physiologicall interpretation.

Electro-physiological recordings usually consist of 
 series of traces, each one corresponds to a sampling
electrode on the tissue. Causality tests are made between the
signals in search for effective connections. 
As Micro Electrode Arrays (MEAs) become denser,the data saturates itself
and the discovery of patterns becomes a difficult task.
Testing for combination of signals in large numbers
brings the so called ``combinatorial explosion'' into
play. 
Novel information is obfuscated by the same high density acquisition that allows us to
discover it. Representing  the recorded data in ways which can unveil previously
invisible patterns is a necessity. Various techniques for separation of functional
generators are available now, most derived from the application of blind source
separation analysis, such as ICA.
In this paper we present a method that aims rather to integrate
the signals into functional components that
have physiologicall interpretation.
This is a technique for tracing out
``functional local connectivity'', which would bring independent information
that is complementary to Blind Source Separation.
Our  method would be applicable mostly to
activity in highly structured neural tissue. We shall exemplify it by
the data obtained from recordings of the rats hippocampus,
taking advantage of its highly stratified  structure.

Local field potential (LFP)
is generated by local multi-unitary activity.
A set of simultaneous and neighbouring traces
can be regarded as representant
of those multi-unitary elements. 
Successive activation  of distinct neighboring elements on structured
tissue would trace a
``trajectory'' of the activity. A tracing of such trajectories would help to indicate
causal or correlated ``functional local connections'' over the tissue. Neighboring
neurons that fire in succession, integrated sub-threshold activity and ephaptic coupling
could all be encompassed as pathways for transmission of activity. Nevertheless,
how to define and acquire the putative paths is an open question.

In previous works some form of vector averaging has been proposed and shown to reveal some trajectories drawn by the activity. We shall call such methods Center of Mass Analyses (CMA).  The ``weights'' that have been chosen for obtaining the average have been selected on heuristic grounds. In the works of Chao \emph{et al.} \cite{Chao05, Chao07}, the density of spikes in a certain time interval was used, a sort of ``averaged density of action'', while Manjarrez \emph{et al.} \cite{Manjarrez07, Manjarrez09} used the positive half of the LFP recording as their weight. 
Here we advance on such methods by incorporating mathematically sound definitions and concepts which allow us to rigorously separate spatially distinct traces. Our weight is actually the natural density function provided by the data: the Current Source Density. This representation not only provides a mathematical rigorous  measure for the weights in the vector averaging process, but helps to separate the data into disjoint sets and disconnected components.
This allows us to track separate process in the same region of interest
(r.o.i.) by following separately the distinct poles of action: influx and
efflux of charged ions.

The application of the method requires electro-physiological data acquired
through bi-dimensional MEAs, although the extension to three dimensional arrays would be possible in principle. The MEA should have dense enough spatial and temporal samplings, so as to obtain good estimates of the CSD and their changes in time.
1
Our method permits us to map possible trajectories of activity inside structured neuronal tissue, without requiring ad hoc geometric constraints. As a prime example, non-convex structures which may show simultaneous activity in separate parts could show the robustness of our method. We have tested then the method on slices of rat's hippocampus, concentrating in the CA3 region. We exemplify the usage of the method with data that has been obtained on experiments on rat's hippocampal slices that are centered on the GD and CA
regions. The slices are put over an array of 64 by 64 electrodes
(BioCAM 4096 by 3Brain \cite{BioCam}),
and data is recorded at sampling rates between 7 and 14 kHz. The richness of the
data permits us to obtain a prime use case for our method. A
 certain amount of down-scaling would be 
possible, but the array of electrodes should be at least 2 dimensional and
a inter-electrode distance of the order of magnitude of a typical pyramidal soma
(around \mum{100}).

@@

After acquiring data from experiment, we pass the data through some Prepossessing
routines. These serve to detect and discard unusable channels,
such as saturated or failing electrodes. The
set of fail-ling electrodes varies from data set to data set,
as they can be produced by bubbles or by saturation after electrical stimulation
or other causes. 
We call this step the elimination of bad channels. We use a simple criterion for this:
if the absolute value of the signal is above certain threshold (which must be
adjusted for evoked vs spontaneous activity) or if the standard deviation is
larger than certain heuristically determined value, the channel is discarded.

@@

The concepts presented here require first obtaining the CSD from the recorded LFP. Any method would present the same utility, we have used two different
approaches. The first method is a  finite difference operator, while the second is the kCSDA presented by Potworowski \emph{et all} \cite{Potworowski2011}, which is an inverse problem approach.  Where both methods are usable, we prefer the former,
as it is faster and requires less assumptions.
We have used kCSDA where
the difference method is not applicable. The results are
independent of the method except for a multiplicative factor which
can be normalized.

Thanks to the density and scales of the BioCAM 4096,
the numerical finite difference method
produces an excellent approximation to the CSD,
if the number of failing channels is low or
they are outside the r.o.i.
In the experiments that were made without electrical stimulation that was the case,
with the order of magnitude of unusable channels in the tens.
A  numerical Laplacian Operator which reduces the cross effect of rectangular grids is the convolution of the data with the following matrix \cite{Lindberg90}:

@@

On previous works it has been argued that numerical difference operators sacrifice all data on the borders of the array. For electrode arrays of lesser density, this could posse a problem, but in our case we renounce to 253 out of 4095 channels (one is grounded), and those are outside the region of interest.
If very few electrodes fail, more sophisticated methods such as iCSDA \cite{Leski2011} and kCSDA \cite{Potworowski2011} prove little advantage in estimating the sources under the ohmic, isotropic and homogeneous assumptions. In case that around 30 or more electrodes
fail, specially if they are on the region of interest, the use of such methods
becomes necessary. The kCSDA method relies on the
assumption that a finite family of convenient functions can span, with acceptable accuracy, a very good approximation to the  CSD at each instant in time.
A general presentation is in the reference \cite{Potworowski2011},we shall recount only the basic implementation as we used it. The
method can be summarized into two steps:
\begin{enumerate}
\item One presents a family of model functions for the CSD and
obtains the LFP that such functions
would generate.
\item Then one proceeds to calculate
  the coefficients for the entire family by projecting the
  experimental data over this ideal LFP model function family.
\end{enumerate}

@@

Then the kCSDA consist of projecting the experimental data (LFP), frame by frame,
over these model functions and obtaining the  coefficients that would spawn an idealized
LFP approximation. These coefficients are the same that would produce the
corresponding CSD. The first operator is the projection operator $K$ of the experimental
LFP (from now on $V(x,y)$) on the space of the idealized LFP (which we call $V^* (x,y)$).
\begin{equation}
  K_{jk}=\sum_{l=1}^{M}b_l(x_j,y_j)b_l(x_k,y_k).
\end{equation}
The sum is done over the set of $M$ electrodes that we want to use.
The other operator is the one that maps the space of $V^*(x,y)$ into the idealized
CSD functions, now called $C^*(x,y)$. 
\begin{equation}
  \tilde{K}_{jk}=\sum_{l=1}^Mb_l(x_j,y_j)\tilde{b}_l(x_k,y_k).
\end{equation}
Then $C^*(x,y)$ can be obtained on the region of interest by:
\begin{equation}
  C^*(x_j,y_j)=\sum_{k,l=1}^M \tilde{K}_{kj} K^{-1}_{ml} V(x_l,y_l)
\end{equation}
Notice how the method incorporates a denoising mechanism, so
no smoothing procedures are needed.
The great advantage of this method is that is possible to obtain a reasonable
approximation to CSD even when the presence of unusable channels is near or inside
the region of interest. The difference method, on the other hand, affects all surrounding
channels to a failing channel, up to 8 of them, so it becomes unusable if enough
saturated electrodes are over the r.o.i.

@@

%% posiblemente quitar esto 
We have performed comparison of both methods over experimental data that has
no failing channels in order to adjust the parameters of the second method. We
have obtained consistent results between them.

@@

Center of Mass Analysis (CMA) has been used previously in order to track putative  ``trajectories'' of the activity or information across neural tissue or even the whole brain \cite{Chao05, Chao07, Manjarrez07, Manjarrez09}. In such works the chosen ``mass'' was an heuristic measure adequate for representing the coarse-grain picture of displacement of activity. These present the problem of justifying the choice made on practical grounds. For calculating a vectorial average, the adequate weight per vector is a density, a positive quantity that measures the concentration of something per space unit, so it is only adequate that we search for a natural density in the electro-physiological experiments. The Current Source Density fits the mathematical and interpretative requirements. We shall denote it in the following as $C$ (following \cite{Potworowski2011}).  This measure is usually presented in arbitrary units, which permits us to interpret it two-fold. One, directly from its definition, as the \emph{differential divergence} of the vectorial electric current density:
\begin{equation}
  C:=\nabla \cdot J
\end{equation}
Due to conservation of charges, this is the same as as the temporal variation of density of charge:
\begin{equation}
  C=-\frac{\partial \rho}{\partial t}
\end{equation}
Current Source Density as a mathematical density (a quasi-probability density function)  allows  to rigorously apply the concepts of vector averaging. This use has been overlooked. This might have occurred because  samplings at lesser spatial resolution did not permit estimations of this quantity with enough precision to make reliable spatial implications, even with the use of sophisticated CSD estimation methods. In our case, we want to use the density as a mathematician would: a function which assigns weights to points in the space of interest and permits us to carry linear operations over them.  The space here is the sampling points of the electrodes at a given instant in time, and the weight, the CSD, is amount of interchange of charged ions from the inside to the outside of cellular membranes, per unit space.  This is an indicator of activity in the neurons, therefore the CSD is also, indirectly, a \emph{``density of local activity''}. Having both
the mathematician and electro-physiologist interpretations presents we can
perform the following analysis with clear readings.

Apparently, CSD would not fit the requirements of non-negativity for its use as a mathematical density, but this is not a problem.  CSD separates the measured activity into three distinct sets, namely, the set of all sources, the set of all sinks and the zero set. The separation is not quantitative but qualitative: even when the signs of electric charge are arbitrary, the distinction between them is not. Therefore, CSD represents on each of the two active sets (sinks and sources) exactly what we need, a unit of concentration of changing charges per unit of space (in this case, per area), and the sign can be ignored, as long as we perform the operations in each set separately. Also, the separation by the zero value set of the CSD is in contrast to the zero value of the LFP, where it has no precise meaning as it only represents a practical reference value. This is acknowledged in most of the literature, where only the scales of the LFP recording are shown, without reference to the sign of the values. In the CSD representation,  the so called zero set (all the points in our space which have a recorded CSD of zero)  has a precise meaning: it is an instantaneous lack of activity and the border between sinks and sources. This shall be exploited further in the section below.

@@

A first visual inspection of the CSD color map at a given instant in time shows that both sources and sinks appear in rather large and significant patches, with typical
cross lengths around \mum{150}. .

These appear to correspond to the expected poles of firing neurons in structured tissue \cite{Buzsaki2012} and inspired us to separate the sink and source  sets into their \emph{disconnected components}. This concept is used in mathematics to denote subsets of a given geometric set that do not touch each other  \cite{Halmos}. Their rough correspondence to poles of firing neurons is most useful in tracking the successive locus of the activity.

To clarify the concept and how we are we using it an example is in order. In figure \ref{disconnectedsets} A we show geometrical sets enclosed by simple curves as color patches, where every color indicates a Set. The set $A$ consists of a single connected component, that means that one can make  path between any two points in the set that consist entirely of points of the same set. In contrast, the set $B$ is made of two disconnected components, labeled $B1$ and $B2$. If one chooses a point in each of the components, one cannot draw a path joining them without leaving the set. But each one of those components is connected, meaning that if we choose the pair of points in one component, then again we can trace a path joining them in the set. So the set $B$ consists of two disconnected components, but each component can be regarded as a connected set on their own. Likewise the set $C$ is made of three disconnected components. In the subfigure \ref{disconnectedsets} Aiii we show the center of mass of the sets. The set $A$ is concave, so, it can have a center of mass outside the set. The
set $B$ has also a center of mass outside the set. It has two disconnected components that are approximately
equal and separated by a long distance (this is not an implication: a set can be disconnected and have
a CM inside the set). For the set $C$ our approach has been used: we treat every disconnected component of the
set as an independent entity and calculate the center of mass for each one. Notice that this doesn't preclude t
he possibility of having a concave subset with a CM that lies outside. But, as we shall see,
as the electrode resolution is finite, the possibility of having highly non compact shapes is unlikely, and they
tend to have a CM inside or near their borders. In the figure \ref{disconnectedsets} B we show this with experimental
data (we are using evoked activity data, as explained in section \ref{sec:evocada}). In graph Bi we
separate the CSD data into the sink and sources set. We take sinks as example. In the graph Bii
we label each disjoint component, obtained by a single pass algorithm. There are 25 components, and
the bigger two are non convex. In the
subfigure Biii we indicate the CM of only those components that are above certain threshold of
integrated intensity (the circle size  is proportional in area to such intensity, there are five CM to be
seen). Each CM gets listed as a probable step in a trajectory.

@@

    Explanation of the geometrical concepts used on our analysis.
    In the Row A of figures
    with an artificial example the concepts of connectedness and convexity.
    The subfigure Ai shows
    three sets, $A$, $B$, and $C$. $A$ fails to be convex because it is concave,
    and $C$ fails to be
    be convex because it is made of two disconnected components.
    In each of them is possible to draw a line
    connecting two points inside the set that is not entirely inside the set.
    In $B$ that is not possible, the set $B$ is convex in a mathematical sense.
    In subfigure Aii we made some modification to show that
    each set can be made of many disconnected components,
    and each one of them can be convex or not.  In Aiii we show the center of
    mass for $A$ and $B$
    taken as a whole, while for $C$ we break the set into each of the disconnected
    components and calculate each CM separately. There is no guarantee that each
    disconnected component would be convex, but as the resolution of the data is finite,
    small components are more likely to
    be convex. In figure B we show everything with a frame of experimental data.
    In Bi we separate the
    data into three sets, Sources, Sinks, and Zeros.
    We discard the Zeros and keep the active ones.
    In Figure Bii we take the Sink Set and separate it into disconnected components,
    they are labeled from one to 25 in this particular example,
    indicated by color
    For each subset we sum over its mass, and 
    considerate for the next step those who are above certain threshold. The
    next step is shown in Biii, where we plot the Center of Mass for each of
    such subsets. Deviations
    from convexity are tolerable,
    as most of the mass is concentrated in the inner electrodes
    of each set.
  }
  \label{disconnectedsets}
\end{figure}

Theoretically, CSD would divide the set of measuring points into three subsets: the set of all sources, the set of all sinks, and the border between them i.e, the set of all measuring points that have a positive CSD  value, the set of those who have a negative CSD value, and the set of those points having exactly the zero value in CSD space.
Due to the finite precision of the measurement,
it is impossible to define a precise zero value.
The measurements produce a function over the grid elements (the electrodes).
Noise between patches corresponding to a single set often makes bridges
between components, making identification of poles inexact.
In order to avoid this we create a ``thick zero set'' of all the recording
points whose CSD values are inside an error interval around zero.
This separates more sharply the components of the set,
represents our ignorance of precise borders and helps
us to simplify the following analysis.

The ``thick zero set'' would help us to separate the sinks and sources sets into
its disconnected components, which would correspond then to clear patches of
definite activity. If each one of these patches could be
interpreted as an active unit,
we may assign to it a ``center of activity'' in order to follow their displacements.
A Center of Mass for each instantaneous disconnected sink/source would provide us
a pragmatic measurement of an ``active locus''. We can then calculate at each time
frame the CM for each one of those components before tracking displacements of activity.

@@

 In order to perform the analysis described here, we summarize the steps.
 \begin{enumerate}
 \item Acquire high density electro-physiological data. The data must be two
   dimensional
   in space. Distance between the recording sites must be of the order of the typical
   size of the neurons involved or less, and such sites must be on a dense enough grid to perform numerical difference operations over them.
   The sampling frequency should be enough to detect sub-threshold activity,
   of at least 3kHz. Also, the data should come from structured tissue.
   Unordered, highly homogeneous tissue, will not yield interpretable results. 
 \item Calculate the CSD from the data. Any method may be used,
   but computer expensive techniques are discouraged if the
   number of channels is above the thousands.
   Common sense and tests are guidelines here.
   We recommend simpler, difference based approaches wherever possible.
   If many electrodes fail near the r.o.i.,
   kCSD is a good alternative if done with
   parallel routines.
   If too many electrodes are failing over the r.o.i.,
   it is better to discard the data. 
 \item Separate the CSD data into three sets: sources, sinks, and border or indiscernible (below error magnitude) data
   which we call the zeros set. Again, sensible criteria have to be defined.
   A rule of thumb would be to use the mean of the peaks of the noise  of the recording as the amplitude for the ``thick zero'' range of values. 
\item Perform disconnected component analysis in the sources and sinks sets, frame by frame.
  A single pass algorithm here is  adequate \cite{Vincent91, Abubaker07}.
\item  Obtain the CM for each component, using eq. \ref{cmparadisj}. Discard
  all CM that are below some significant threshold value. A rule of thumb is one tenth of the average value for all CM,
  or those who are in the lesser decil. Anything that is around the peak
  values of the noise integrated over around 5 electrodes would have to be discarded.
\item For each frame and its successor, and for each CM, perform a ``most probable successor'' detection. Here we must put numerical criteria according to the data available, i.e. how far away and how different in intensity should be allowed a CM to be at time $t+1$ to be considered successor in time of another at time $t$.
\item ``Connect the dots'', that is, apply successively the step above for each CM, until it stops having a successor or its intensity falls below the error criteria. Then plot the successive CMs as one trajectory.
\item Plot all the trajectories and sort visually those that may have physiological interpretation from those who may be artifacts of the method. This may need some form of expert validation.
 \end{enumerate}

>>>>

Tracking the propagation of activity in neural tissue poses challenging
problems.

Electro-physiological recordings usually consist of 
 series of traces, each one corresponding to a sampling
electrode on the tissue. Causality tests are made between the
signals in search for effective connections. 
As Micro Electrode Arrays (MEAs) become denser,the data saturates itself
and the discovery of patterns becomes a difficult task.
Testing for combination of signals in large numbers
brings ``combinatorial explosion'' into
play. This makes the causality tests time consuming and
prone to produce false positives.
Novel information is obfuscated by the same high density acquisition that allows us to
discover it. Representing  the recorded data in ways which can unveil previously
invisible patterns is a necessity. Various techniques for separation of signals
are available now, most derived from the application of blind source
separation analysis, such as ICA.
In this paper we present a method that aims rather to integrate
the signals into functional components that
have physiologicall interpretation.
This is a technique for tracing out
``functional local connectivity'', which would bring independent information
that is complementary to Blind Source Separation.
Our  method would be applicable mostly to
activity in highly structured neural tissue. We shall exemplify it by
the data obtained from recordings of the rats hippocampus,
taking advantage of its highly stratified  structure.

Local field potential (LFP)
is generated by local multi-unitary activity.
A set of simultaneous and neighbouring traces
can be regarded as representant of the action
of those multi-unitary elements. 
Successive activation  of distinct neighboring elements on structured
tissue would trace a
``trajectory'' of the activity. A tracing of such trajectories would help to indicate
causal or correlated ``functional local connections'' over the tissue. Neighboring
neurons that fire in succession, integrated sub-threshold activity and ephaptic coupling
could all be encompassed as pathways for transmission of activity. Nevertheless,
how to define and acquire the putative paths is an open question.

In previous works some form of vector averaging has been proposed and shown to reveal some trajectories drawn by the activity. We shall call such methods Center of Mass Analyses (CMA).  The ``weights'' that have been chosen for obtaining the average have been selected on heuristic grounds. In the works of Chao \emph{et al.} \cite{Chao05, Chao07}, the density of spikes in a certain time interval was used, a sort of ``averaged density of action'', while Manjarrez \emph{et al.} \cite{Manjarrez07, Manjarrez09} used the positive half of the LFP recording as their weight. 
Here we advance on such methods by incorporating mathematically sound definitions and concepts which allow us to rigorously separate spatially distinct traces. Our weight is actually the natural density function provided by the data: the Current Source Density. This representation not only provides a mathematical rigorous  measure for the weights in the vector averaging process, but helps to separate the data into disjoint sets and disconnected components.
This allows us to track separate process in the same region of interest
(r.o.i.) by following separately the distinct poles of action: influx and
efflux of charged ions, which correspond to two distinct phases of the
neuronal activity: depolarization and repolarization.

The application of the method requires electro-physiological data acquired
through bi-dimensional MEAs, although the extension to three dimensional arrays
is straightforward.
The MEA should have dense enough spatial and temporal samplings,
so as to obtain good estimates of the CSD and their changes in time.

Our method permits us to map possible trajectories of activity inside structured neuronal tissue, without requiring ad hoc geometric constraints. As a prime example, non-convex structures which may show simultaneous activity in separate parts could show the robustness of our method. We have tested the method on slices of rat's hippocampus, concentrating in the CA3 region. We exemplify the usage of the method with data that has been obtained on experiments on rat's hippocampal slices that are centered on the GD and CA
regions. The slices are put over an array of 64 by 64 electrodes
(BioCAM 4096 by 3Brain \cite{BioCam}),
and data is recorded at sampling rates between 7 and 14 kHz. The richness of the
data is an exemplary use case for our method. A certain amount of down-scaling would be 
possible, but the array of electrodes should be at least 2 dimensional and
a inter-electrode distance of the order of magnitude of a typical pyramidal soma
(around \mum{100} at most).

@@

After acquiring data from experiment, we pass the data through some Prepossessing
routines. These serve to detect and discard unusable channels,
such as saturated or failing electrodes. The
set of failling electrodes varies from data set to data set,
as they can be produced by bubbles or by saturation after electrical stimulation
or any other cause. 
We call this step the elimination of bad channels. We use a simple criterion for this:
if the absolute value of the signal is above certain threshold (which must be
adjusted for evoked vs spontaneous activity, but
is of the order of \muV{1000}) or if the standard deviation is
larger than a two times the averague r.m.s.  of the normal noise,
the channel is discarded.

@@

The first step is to obtain the CSD from the recorded LFP.
Any reliable method would work.
We have used two different
approaches. The first method is a  finite difference operator,
while the second is the kCSDA presented by
Potworowski \emph{et all} \cite{Potworowski2011},
which is an inverse problem approach.
Where both methods are usable, we prefer the former,
as it is faster and requires less assumptions.
We have used kCSDA where
the difference method is not applicable. The results are
independent of the method except for a multiplicative factor.
As there is still another unknown constant factor
(the resistivity of the medium) this is irrelevant.
Therefore, all of the CSD results are presented in arbitrary units.

Thanks to the density and scales of the BioCAM 4096,
the numerical finite difference method
produces an excellent approximation to the CSD,
if the number of failing channels is low or
they are outside the r.o.i.
In most examples the number  of unusable channels is in the tens.
A  numerical Laplacian Operator which reduces the cross effect of
rectangular grids is the convolution of the data with the following
matrix \cite{Lindberg90}:

@@

On previous works it has been argued that numerical difference operators sacrifice all data on the borders of the array. For electrode arrays of lesser density, this could posse a problem, but in our case we renounce to 253 out of 4095 channels (one is grounded), and those are outside the region of interest.
If very few electrodes fail, more sophisticated methods such as iCSDA \cite{Leski2011} and kCSDA \cite{Potworowski2011} prove little advantage.
In case that around 30 or more electrodes
fail, specially if they are on the region of interest or its borders,
the use of such methods
becomes necessary. The kCSDA method relies on the
assumption that a finite family of convenient functions can span, with acceptable accuracy, a very good approximation to the  CSD at each instant in time.
A general presentation is in the reference \cite{Potworowski2011},we shall recount only the basic implementation as we used it. The
method can be summarized into two steps:
\begin{enumerate}
\item One presents a family of model functions for the CSD and
obtains the LFP that such functions
would generate.
\item Then one proceeds to calculate
  the coefficients for the entire family by projecting the
  experimental data over this ideal LFP model function family.
\end{enumerate}

@@

Then the kCSDA consist of projecting the experimental data (LFP), frame by frame,
over these model functions and obtaining the  coefficients that would spawn an idealized
LFP approximation. These coefficients are the same that would produce the
corresponding CSD. The first operator is the projection operator $K$ of the experimental
LFP (from now on $V(x,y)$) on the space of the idealized LFP (which we call $V^* (x,y)$).
\begin{equation}
  K_{jk}=\sum_{l=1}^{M}b_l(x_j,y_j)b_l(x_k,y_k).
\end{equation}
The sum is done over the set of $M$ electrodes that we want to use.
The other operator is the one that maps the space of $V^*(x,y)$ into the idealized
CSD functions, now called $C^*(x,y)$. 
\begin{equation}
  \tilde{K}_{jk}=\sum_{l=1}^Mb_l(x_j,y_j)\tilde{b}_l(x_k,y_k).
\end{equation}
Then $C^*(x,y)$ can be obtained on the region of interest by:
\begin{equation}
  C^*(x_j,y_j)=\sum_{k,l=1}^M \tilde{K}_{kj} K^{-1}_{ml} V(x_l,y_l)
\end{equation}
Notice how the method incorporates a denoising mechanism, so
no smoothing filters are needed.
The great advantage of this method is that is possible to obtain a reasonable
approximation to CSD even when the presence of unusable channels is near or inside
the region of interest. The difference method, on the other hand, affects all surrounding
channels to a failing channel, up to 8 of them, so it becomes unusable if enough
saturated electrodes are over the r.o.i.

@@

@@

Center of Mass Analysis (CMA) has been used previously in order to track putative  ``trajectories'' of the activity or information across neural tissue or even the whole brain \cite{Chao05, Chao07, Manjarrez07, Manjarrez09}. In such works the chosen ``mass'' was an heuristic measure adequate for representing the coarse-grain picture of displacement of activity. These present the problem of justifying the choice made on practical grounds. For calculating a vectorial average, the adequate weight per vector is a density, a positive quantity that measures the concentration of something per space unit, so it is only adequate that we search for a natural density in the electro-physiological experiments. The Current Source Density fits the mathematical and interpretative requirements. We shall denote it in the following as $C$ (following \cite{Potworowski2011}).  This measure is usually presented in arbitrary units, which permits us to interpret it two-fold. One, directly from its definition, as the \emph{differential divergence} of the vectorial electric current density:
\begin{equation}
  C:=\nabla \cdot J
\end{equation}
Due to conservation of charges, this is the same as as the temporal variation of density of charge:
\begin{equation}
  C=-\frac{\partial \rho}{\partial t}
\end{equation}
Current Source Density as a mathematical density (a quasi-probability density function)  allows  to rigorously apply the concepts of vector averaging. This use has been overlooked. This might have occurred because  samplings at lesser spatial resolution did not permit estimations of this quantity with enough precision to make reliable spatial implications, even with the use of sophisticated CSD estimation methods. In our case, we want to use the density as a mathematician would: a function which assigns weights to points in the space of interest and permits us to carry linear operations over them.  The space here is the sampling points of the electrodes at a given instant in time, and the weight, the CSD, is amount of interchange of charged ions from the inside to the outside of cellular membranes, per unit space.  This is an indicator of activity in the neurons, therefore the CSD is also, indirectly, a \emph{``density of local activity''}. Having both
the mathematician and electro-physiologist interpretations present we can
perform the following analysis.

Apparently, CSD would not fit the requirements of non-negativity for its use as a mathematical density. This is not a problem.  CSD separates the measured activity into three distinct sets, namely, the set of all sources, the set of all sinks and the zero set. The separation is not quantitative but qualitative: even when the signs of electric charge are arbitrary, the distinction between them is not. Therefore, CSD represents on each of the two active sets (sinks and sources) exactly what we need, a unit of concentration of changing charges per unit of space (in this case, per area), and the sign can be ignored, as long as we perform the operations in each set separately. Also, the separation by the zero value set of the CSD is in contrast to the zero value of the LFP, where it has no precise meaning as it only represents a practical reference value. This is acknowledged in most of the literature, where only the scales of the LFP recording are shown, without reference to the sign of the values. In the CSD representation,  the so called zero set (all the points in our space which have a recorded CSD of zero)  has a precise meaning: it is an instantaneous lack of activity and the border between sinks and sources. This shall be exploited further in the section below.

@@

A first visual inspection of the CSD color map at a given instant in time shows that both sources and sinks appear in rather large and significant patches, with typical
cross lengths around \mum{150}. .
These appear to correspond to the expected poles of firing neurons in structured tissue \cite{Buzsaki2012} and inspired us to separate the sink and source  sets into their \emph{disconnected components}. This concept is used in mathematics to denote subsets of a given geometric set that do not touch each other  \cite{Halmos}. Their rough correspondence to poles of firing neurons is most useful in tracking the successive locus of the activity.

To clarify the concept and how we are we using it an example is in order. In figure \ref{disconnectedsets} \textbf{A} we show geometrical sets enclosed by simple curves as color patches, where every color indicates a Set. The set $A$ consists of a single connected component, that means that one can make  path between any two points in the set that consist entirely of points of the same set. In contrast, the set $B$ is made of two disconnected components, labeled $B1$ and $B2$. If one chooses a point in each of the components, one cannot draw a path joining them without leaving the set. But each one of those components is connected, meaning that if we choose the pair of points in one component, then again we can trace a path joining them in the set. So the set $B$ consists of two disconnected components, but each component can be regarded as a connected set on their own. Likewise the set $C$ is made of three disconnected components. In the subfigure \ref{disconnectedsets} \textbf{Aiii} we show the center of mass of the sets. The set $A$ is concave, so, it can have a center of mass outside the set. The
set $B$ has also a center of mass outside the set. It has two disconnected components that are approximately
equal and separated by a long distance (this is not an implication: a set can be disconnected and have
a CM inside the set). For the set $C$ our approach has been used: we treat every disconnected component of the
set as an independent entity and calculate the center of mass for each one. Notice that this doesn't preclude t
he possibility of having a concave subset with a CM that lies outside. But, as we shall see,
as the electrode resolution is finite, the possibility of having highly non compact shapes is unlikely, and they
tend to have a CM inside or near their borders. In the figure \ref{disconnectedsets}
\texxtbf{B} we show this with experimental
data (we are using evoked activity data, as explained in section \ref{sec:evocada}). In graph
\textbf{Bi} we
separate the CSD data into the sink and sources set. We take sinks as example. In the graph
\textbf{Bii}
we label each disjoint component, obtained by a single pass algorithm. There are 25 components, and
the bigger two are non convex. In the
subfigure \textbf{Biii}
we indicate the CM of only those components that are above certain threshold of
integrated intensity (the circle size  is proportional in area to such intensity, there are five CM to be
seen). Each CM gets listed as a probable step in a trajectory.

@@

    Explanation of the geometrical concepts used on our analysis.
    In the Row \textbf{A} of figures we show
    with an artificial example the concepts of connectedness and convexity.
    The subfigure \textbf{Ai} shows
    three sets, $A$, $B$, and $C$. $A$ fails to be convex because it is concave,
    and $C$ fails to be
    be convex because it is made of two disconnected components.
    In each of them is possible to draw a line
    connecting two points inside the set that is not entirely inside the set.
    In $B$ that is not possible, the set $B$ is convex in a mathematical sense.
    In subfigure \textbf{Aii} we made some modification to show that
    each set can be made of many disconnected components,
    and each one of them can be convex or not.  In \textbf{Aiii} we show the center of
    mass for $A$ and $B$
    taken as a whole, while for $C$ we break the set into each of the disconnected
    components and calculate each CM separately. There is no guarantee that each
    disconnected component would be convex, but as the resolution of the data is finite,
    small components are more likely to
    be convex. In figure \textbf{B} we show everything with a frame of experimental data.
    In \textbf{Bi} we separate the
    data into three sets, Sources, Sinks, and Zeros.
    We discard the Zeros and keep the active ones.
    In Figure \textbf{Bii}
    we take the Sink Set and separate it into disconnected components,
    they are labeled from one to 25 in this particular example,
    indicated by color
    For each subset we sum over its mass, and 
    considerate for the next step those who are above certain threshold. The
    next step is shown in \textbf{Biii},
    where we plot the Center of Mass for each of
    such subsets. Deviations from convexity are tolerable,
    as most of the mass is concentrated in the inner electrodes
    of each set.
  }
  \label{disconnectedsets}
\end{figure}

Theoretically, CSD would divide the set of measuring points into three subsets: the set of all sources, the set of all sinks, and the border between them.
It is impossible to define a precise zero value.
The measurements produce a function over the grid elements (the electrodes).
Noise between patches corresponding to a single set often makes bridges
between components, making identification of poles inexact.
In order to avoid this we create a ``thick neutral set'' of all the recording
points whose CSD values are inside an error interval around zero.
This separates more sharply the components of the set,
represents our ignorance of precise borders and helps
us to simplify the following analysis.

The ``thick neutral set'' would help us to separate the sinks and sources sets into
its disconnected components, which would correspond then to clear patches of
definite activity. If each one of these patches could be
interpreted as an active unit,
we may assign to it a ``center of activity'' in order to follow their displacements.
A Center of Mass for each instantaneous disconnected sink/source would provide us
a pragmatic measurement of an ``active locus''. We can then calculate at each time
frame the CM for each one of those components before tracking displacements of activity.

@@

 In order to perform the analysis described here, we summarize the steps.
 \begin{enumerate}
 \item Acquire high density electro-physiological data. The data must be two
   dimensional
   in space. Distance between the recording sites must be of the order of the typical
   size of the neurons involved or less, and such sites must be on a dense enough grid to perform numerical difference operations over them.
   The sampling frequency should be enough to detect sub-threshold activity,
   of at least 3kHz. Also, the data should come from structured tissue.
   Unordered, highly homogeneous tissue, will not yield interpretable results. 
 \item Calculate the CSD from the data. Any method may be used,
   but computer expensive techniques are discouraged if the
   number of channels is above the thousands.
   Common sense and tests are guidelines here.
   We recommend simpler, difference based approaches wherever possible.
   If many electrodes fail near the r.o.i.,
   kCSD is a good alternative if performed using
   parallel CPU or GPU computing.
   If too many electrodes are failing over the r.o.i.,
   it is better to discard the data. 
 \item Separate the CSD data into three sets: sources, sinks, and neutral.
   Again, sensible criteria have to be defined.
   A rule of thumb would be to use the mean of the peaks of the noise  of the recording as the amplitude for the ``thick neutral'' range of values. 
\item Perform disconnected component analysis in the sources and sinks sets, frame by frame.
  A single pass algorithm here is  adequate \cite{Vincent91, Abubaker07}.
\item  Obtain the CM for each component, using eq. \ref{cmparadisj}. Discard
  all CM that are below some significant threshold value. A rule of thumb is one tenth of the average value for all CM,
  or those who are in the lesser decil. 
\item For each frame and its successor, and for each CM, perform a ``most probable successor'' detection. Here we must put numerical criteria according to the data available, i.e. how far away and how different in intensity should be allowed a CM to be at time $t+1$ to be considered successor in time of another at time $t$.
\item ``Connect the dots'', that is, apply successively the step above for each CM, until it stops having a successor or its intensity falls below the error criteria. Then plot the successive CMs as one trajectory.
\item Plot all the trajectories and sort visually those that may have physiological interpretation from those who may be artifacts of the method. This may need some form of expert validation.
 \end{enumerate}
